This dissertation presents the design, implementation and evaluation of a multiclient file server for the seL4 Microkit framework, motivated by the aim of mitigating Inter-Process Communication (IPC) overhead — a key performance constraint in microkernel-based systems — through asynchronous client-server communication. The system supports a rich set of file system operations, including namespace, data and metadata management, and multiclient access control.

The file server is structured around flexible and modular core abstractions, including blocks, inodes, directories and file descriptors, and interfaced by two communication models: a synchronous model based on protected procedure calls and an asynchronous model based on notifications, shared memory queues and \textit{batched} request processing, where multiple operations may be sent and processed collectively. A transparent Application Programming Interface (API) was implemented to abstract these communication details while preserving consistent operation semantics across both models.

The asynchronous model demonstrated reduced IPC overhead through batching operations and reducing synchronisation frequency, improving throughput and lowering average operation latency. However, these benefits proved workload-dependent, with limited gains under highly interdependent workloads. A key insight is that communication structure alone can significantly influence system behaviour independently from core file system design. The results further suggest that, although evaluated in a single-threaded environment, the asynchronous design could enable substantially greater concurrent processing on multicore systems through non-blocking execution and overlap of request generation, servicing and response handling.

This work demonstrates that complex, stateful services can be realised entirely in user space within the Microkit framework for seL4 and provides insight into trade-offs between synchronous and asynchronous communication for future microkernel-based system services.
